How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2024-09-11 1.10-lifescientific.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

Throughout the book, I make reference to the ‘scientific project’ as a greater plan to which our studies in biological sciences is simply a part. I consider the scientific project as a path that humanity should continue to travel in order to reach a higher and more just way of living. Our societies have evolved from places where superstition and belief were the basis for societal rules that favoured the privileged classes. Science is a process in which we reach conclusions based on egalitarian reasoning and evidence, something that we can aspire for the larger community around us. Already, most societies are making the majority of their collective evidence-based decisions through the guidance of science. Therefore, if we consider our knowledge-based societies to be democratic, then all citizens of that society should understand the processes under which the foundation of that knowledge is acquired [@sagan2011demon]. Yet globally, public skepticism in science has been rising, and has reached as high as 35%. To consolidate this growing position of science in society, and the decline in public acceptance, scientists will need to become more open, communicative and seen to be conducting their own profession in a way that reflects these ideals. There is evidence that increasing communication from scientists to the public results in a decrease of skepticism ([see here](https://multimedia.3m.com/mws/media/1898512O/3m-sosi-2020-pandemic-pulse-global-report-pdf.pdf)).

@@

## Should society lead the sciences or sciences lead society? 
Of course, the answer is simply that there must be feedback between science and society. As science advances, society learns of these advances and views are advanced. To become incorporated as a societal norm takes time. Over time, views from science that appear radical to society become mainstream and accepted. Think of the views on the earth being round, or more recently, the general acceptance that the universe started with the ‘big bang’. The counterculture, also known as post-truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon. Indeed, in the scientific project, we need to accept that these views will persist and empower the general populace to understand why and how they can be shown to be false. Hence, your view that the earth is not flat is bolstered if you can provide three ways to demonstrate this as fact (pick the ones you like from 7 listed [here](http://crosstalk.cell.com/blog/seven-ways-to-prove-earth-is-round) or [here](https://www.unlv.edu/news/release/round-earth-clues-how-science-proves-our-home-globe)). Otherwise, your assertion that it is not flat will appear as easily denied as a ‘flat-earther’s’ assertion that it is. Thus the power of the scientific argument in society is that it is not simply contradiction or gainsay but that there is evidence.

The scientific project is currently not the open virtuous eutopia that we would like. Indeed, as scientists are humans and part of the current society, they are prone to the same trappings of wielding power and influence in order to get ahead. As you get toward the end of this book, you will see me calling out increasing parts of the scientific project that are currently riddled with bias, thereby weakening the entire process and disenfranchising the most vulnerable. Reforming science is no easy task. Reforms are needed throughout the structures that fund, employ and disseminate scientific research. Sweeping this all away will take a concerted effort but might well be easier in science than in other sectors of society. Nearly all scientists recognise that science needs to be objective and impartial in order to succeed. Their objectivity makes them an ideal group to kickstart a fairer, more transparent and open system. As you will see, there has already been a move to make science more open, and by embracing this transparency the next generation of scientists will reap the benefits, and be on a sounder footing to improve a better scientific society.

@@

One aspect of this will be to bring more of the population into conducting scientific studies. This can be at the level of funding scientific studies. Crowdsourced scientific studies promote communication between the scientists and the funders to a level that is meaningful to both groups, such that they make up a large scale collaborative research interaction [@uhlmann2019scientific]. Citizen science contributions mean that the collection of data can be shared on a much broader scale, than would be traditionally possible. Of particular interest in the biological sciences is the identification and collection of locality data for biological species (on platforms such as iNaturalist). But increasing numbers of projects are demonstrating novel ways for the public to increasingly participate meaningfully in scientific endeavour [@bonney2014next; @silvertown2009new]. Science is a joyful and spiritual experience for most of us that participate in it. There’s no reason not to spread that joy as widely as possible.

>>>>

Throughout the book, I refer to the ‘scientific project’ as a greater plan of which our studies in biological sciences are simply a part. I consider the scientific project as a path that humanity should continue to travel to reach a higher and more just way of living. Our societies have evolved from places where superstition and belief were the basis for societal rules that favoured the privileged classes. Science is a process in which we reach conclusions based on egalitarian reasoning and evidence, something that we can aspire for in the larger community around us. Already, most societies are making the majority of their collective evidence-based decisions through the guidance of science. Therefore, if we consider our knowledge-based societies to be democratic, then all citizens of that society should understand the processes under which the foundation of that knowledge is acquired [@sagan2011demon]. Yet globally, public skepticism in science has been rising and has reached as high as 35% ([see here](https://multimedia.3m.com/mws/media/1898512O/3m-sosi-2020-pandemic-pulse-global-report-pdf.pdf)). To consolidate this growing position of science in society, and the decline in public acceptance, scientists will need to become more open, communicative and seen to be conducting their profession in a way that reflects these ideals. There is evidence that increasing communication from scientists to the public results in a decrease in scepticism [@goldenberg2023public].

@@

## Should society lead the sciences or sciences lead society? 
Of course, the answer is simply that there must be feedback between science and society. As science advances, society learns of these advances and views are advanced. To become incorporated as a societal norm takes time. Over time, views from science that appear radical to society become mainstream and accepted. Think of the views on the earth being round, or more recently, the general acceptance that the universe started with the ‘big bang’. The counterculture faction in society, also known as post-truth, that espouses views accepted by the mainstream and proved by science beyond a reasonable doubt is not a novel phenomenon. Indeed, in the scientific project, we need to accept that these views will persist and empower the general populace to understand why and how they can be shown to be false. Hence, your view that the earth is not flat is bolstered if you can provide three ways to demonstrate this as fact (pick the ones you like from 7 listed [here](http://crosstalk.cell.com/blog/seven-ways-to-prove-earth-is-round) or [here](https://www.unlv.edu/news/release/round-earth-clues-how-science-proves-our-home-globe)). Otherwise, your assertion that it is not flat will appear as easily denied as a ‘flat-earther’s’ assertion that it is. Thus the power of the scientific argument in society is that it is not simply contradiction or gainsay but that there is evidence.

The scientific project is currently not the open virtuous eutopia that we would like. Indeed, as scientists are humans and part of the current society, they are prone to the same trappings of wielding power and influence to get ahead. As you get toward the end of this book, you will see me calling out increasing parts of the scientific project that are currently riddled with bias, thereby weakening the entire process and disenfranchising the most vulnerable. Reforming science is no easy task. Reforms are needed throughout the structures that fund, employ and disseminate scientific research. Sweeping this all away will take a concerted effort but might well be easier in science than in other sectors of society. Nearly all scientists recognise that science needs to be objective and impartial in order to succeed. Their objectivity makes them an ideal group to kickstart a fairer, more transparent and open system. As you will see, there has already been a move to make science more open, and by embracing this transparency the next generation of scientists will reap the benefits, and be on a sounder footing to improve a better scientific society.

@@

One aspect of this will be to bring more of the population into conducting scientific studies. This can be at the level of funding scientific studies. Crowdsourced scientific studies promote communication between the scientists and the funders to a level that is meaningful to both groups, such that they make up a large scale collaborative research interaction [@uhlmann2019scientific]. Citizen science contributions mean that the collection of data can be shared on a much broader scale, than would be traditionally possible. Of particular interest in the biological sciences is the identification and collection of locality data for biological species (on platforms such as [iNaturalist](www.inaturalist.org)). But increasing numbers of projects are demonstrating novel ways for the public to increasingly participate meaningfully in scientific endeavour [@bonney2014next; @silvertown2009new]. Science is a joyful and spiritual experience for most of us that participate in it. There’s no reason not to spread that joy as widely as possible.

### 2024-09-11 2.03-NoHyp.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

1. Observational science (something that I refer to elsewhere as [Natural History Observations](https://howtopublishscience.org/whattowrite.html#natural-history-observations)) can teach us a lot more about systems and inspire greater creativity in our work than plodding from experiment to experiment. 
1. We can infer patterns from process. As humans, we are evolutionarily programmed to see patterns (even when they aren't there) and recognise patterns in the world around us [e.g., @linhares2013what]. It would be crazy to put this ability aside and ignore it when conducting science. Clearly, we are also able to make mistakes when we infer patterns from processes, but this does not mean that it should not be done. But we need to engage in a much more cautious manor when using this approach, always being aware of our bias towards certain explanations so that we encompass others. I would also argue here that this is instrumental in forming hypotheses to test experimentally or in models. 
1. Models of systems are powerful tools to test questions. Models are always simplified versions of complex biological systems, but their power as tools in our scientific approach should no longer be questioned.

>>>>

1. Observational science (something that I refer to elsewhere as [Natural History Observations](https://howtopublishscience.org/whattowrite.html#natural-history-observations)) can teach us a lot more about systems and inspire greater creativity in our work than plodding from experiment to experiment. 
1. We can infer patterns from process. As humans, we are evolutionarily programmed to see patterns (even when they aren't there) and recognise patterns in the world around us [e.g., @linhares2013what]. It would be crazy to put this ability aside and ignore it when conducting science. Clearly, we are also able to make mistakes when we infer patterns from processes, but this does not mean that it should not be done. But we need to engage in a much more cautious manner when using this approach, always being aware of our bias towards certain explanations so that we encompass others. I would also argue here that this is instrumental in forming hypotheses to test experimentally or in models. 
1. Models of systems are powerful tools to test questions. Models are always simplified versions of complex biological systems, but their power as tools in our scientific approach should no longer be questioned.

### 2024-09-11 2.05a-AI.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

The emergence of artificial intelligence (AI) platforms brings with it new issues around writing generally, including writing your PhD thesis. These bots are already widely available in early 2023, with both Microsoft (through their company OpenAI) and Google launching [ChatGPT](chat.openai.com) and [Bard](bard.google.com), respectively. There are already publications that list ChatGPT as an author [@oconnor2023open; @mijwil2023artificial] in acknowledgement that a large amount of the text therein had been generated by the bot. More AI writing tools are going to be coming online in the near future. Microsoft's plan to integrate the ChatGPT into Bing and the Bing platform directly into their office suite, means that as an author you may unwittingly use AI generated automated suggestions inside your word processor without realising it. Thus, you should be interested in knowing whether such tools can be used to help you in your own writing. If you haven't already tried [@owens2023how], once you do you will understand the tantalising possibility that they could be used to replace your needing to write.

@@

@@

**ChatGPT**

@@

**Bard**

@@

To summarise, LLMs are trained on an exceptionally large amount of data and use AI to first understand what you are asking it, and second pull out of their training set what they need to reply to your user input. The key then is that whereas previous chat bots had (relatively) small training sets and limited ability to understand user input, these new LLMs have vast amount of data.

@@

### Plagiarism by AI
There is the chance that your AI tool is going to plagiarise text that it finds on the internet, and so you may unwittingly be pasting the words of another author into your own text (see more on plagiarism [here](#plagiarise)). Some authors consider that even if the AI is programmed to avoid plagiarism, using this text in any way should be considered as plagiarism [@salvagno2023can]. However, on balance, my own view is that any attempt to be overly restrictive on the use of AI generated text is bound to fail as AI itself becomes increasingly more sophisticated and widespread. Instead, I share the view [@lim2023generative] that it is better to embrace and learn how to use AI as a tool. Learn about what it can do, as well as its limitations.

@@

Clearly, the definition of plagiarism will need some work in the light of new abilities for AIs to write text.

@@

Plagiarism is the subject of another chapter, and you can read more about it [there](#plagiarise).

@@

### Use AI as a tool 
Essentially, if you want to try using AI to help you write, use it as a tool. Don't try to use it to replace your own writing. That will likely end in a poor product, and could possibly end with factually incorrect or biased text that you won't want to be associated with your name - and clearly not something that you should want your advisor to read.

Getting started is often a struggle and I regularly suggest that students start by writing something/anything to fill out their outline, even if all of that text gets replaced at a later date. This is a way that you could use a tool like ChatGPT. Produce your own outline, because you do still need to plan your chapter, but then try using an AI like ChatGPT to flesh this out. Hopefully, this will end by helping you to write your own text, replacing everything that the AI has produced.

It is worth noting that the input for ChatGPT is dated to a portion of the internet from September 2021 [@azamfirei2023large]. It cannot access the internet and so you will not get anything more recent. Whereas Bard is trained from Google search results, Wikipedia, [The Common Crawl](commoncrawl.org) and other publicly available datasets.

@@

## Will others detect ChatGPT if I use it?
Some people claim that they can detect the output from ChatGPT as it lacks the depth and insight that original authors usually have. In other words, the aim of ChatGPT is to produce the words (in a grammatically correct manor), while the aim of an author is to transmit an idea to the reader. As you might expect and hope, AI is not at the point where it can generate the same intent to communicate.

>>>>

The emergence of artificial intelligence (AI) platforms brings with it new issues around writing generally, including writing your PhD thesis. These bots are already widely available in early 2023, with both Microsoft (through their company OpenAI) and Google launching [ChatGPT](chat.openai.com) and [Gemini](gemini.google.com), respectively. Note that Google originally launched as Bard in early 2023, but then later released as an update known as Gemini. There are already publications that list ChatGPT as an author [@oconnor2023open; @mijwil2023artificial] in acknowledgement that a large amount of the text therein had been generated by the bot. More AI writing tools are going to be coming online in the near future. Microsoft's plan to integrate the ChatGPT into Bing and the Bing platform directly into their office suite, means that as an author you may unwittingly use AI generated automated suggestions inside your word processor without realising it. Thus, you should be interested in knowing whether such tools can be used to help you in your own writing. If you haven't already tried [@owens2023how], once you do you will understand the tantalising possibility that they could be used to replace your needing to write.

@@

## This field is moving very fast {.unlisted .unnumbered}

Bear in mind when you read this that the original chapter was written in March 2023, with updates nearly every month since then. This current version was compiled on `r format(Sys.time(), '%d %B, %Y')`. If you think that the content here (or elsewhere in the book) is out of date, or needs additional text, please let me know (see [here](#contribute)).

@@

#### **ChatGPT** (March 2023) {.unlisted .unnumbered}

@@

#### **Bard** (March 2023) {.unlisted .unnumbered}

@@

#### **Gemini** (May 2024) - note the shift from using first person (Bard) to third person (Gemini). It's a deliberate change made by Google. {.unlisted .unnumbered}

>Gemini is actually the new name for what was formerly known as Bard. It's like a more powerful and upgraded version. Here's the breakdown of the key differences:

>Underlying Tech: Bard primarily relied on Recurrent Neural Networks (RNNs) for processing information. Gemini utilizes a transformer-based design, which is considered more powerful for various tasks.
Training Data: Gemini is trained on a much larger and more diverse dataset compared to Bard. This includes text, code, and even images, allowing it to understand and respond to a wider range of prompts and questions.
Capabilities: Bard was focused on text-based interactions. Gemini, however, has multimodal capabilities. This means it can not only process text but also understand and generate different formats of data, including images and potentially even video.
In short, Gemini is designed to be more versatile and powerful than Bard,  with the ability to handle complex tasks and integrate with other Google tools more seamlessly.

>To summarise, LLMs are trained on an exceptionally large amount of data and use AI to first understand what you are asking it, and second pull out of their training set what they need to reply to your user input. The key then is that whereas previous chat bots had (relatively) small training sets and limited ability to understand user input, these new LLMs have vast amount of data.

@@

### Plagiarism by AI
There is the chance that your AI tool is going to plagiarise text that it finds on the internet, and so you may unwittingly be pasting the words of another author into your own text (see more on plagiarism [here](#plagiarise)). Some authors consider that even if the AI is programmed to avoid plagiarism, using this text in any way should be considered as plagiarism [@salvagno2023can]. Certainly, if you use AI generated text and then claim that you have written this yourself, then this is clearly plagiarism. In this case, the plagiarism is due to the false claim that you wrote the text.

@@

Clearly, the definition of plagiarism will need some work in the light of new abilities for AIs to write text. What I see happening is that both universities and publishers are asking students / authors to declare that they have not generated any text using AI. This is a catch-all and will mean that you simply cannot use AI generated text.

@@

My own view is that any attempt to be overly restrictive on the use of AI generated text is bound to fail as AI itself becomes increasingly more sophisticated and widespread. Instead, I share the view of Lim [-@lim2023generative] that it is better to embrace and learn how to use AI as a tool. Learn about what it can do, as well as its limitations.
There is an entire chapter on plagiarism (in [Part II](#plagiarise)) where AI generated text is covered, and I'd advise you to read this to learn more.

@@

### Use AI as a tool 
Essentially, if you want to try using AI to help you write, use it as a tool. Don't try to use it to replace your own writing. That will likely end in a poor product, and could possibly end with factually incorrect or biased text that you won't want to be associated with your name, and clearly not something that you should want your advisor to read.

Getting started is often a struggle and I regularly suggest that students start by writing something/anything to fill out their outline, even if all of that text gets replaced at a later date. This is a way that you could use AI as a tool. Produce your own outline, because you do still need to plan your chapter, but then try using an AI to flesh this out. Read through the output and get inspired. Hopefully, this will end by helping you to write your own text, replacing everything that the AI has produced.

#### **But do not copy and paste it into your document.** {.unlisted .unnumbered}

Use the AI to help you get stated but don't copy it. Just like you might read the text of an article and that helps you get some understanding, you can use AI generated text in the same way. But if you copy and paste it, you'll start on a slippery slope towards plagiarism and potential disqualification.

#### **Be aware that AI can get it wrong.** {.unlisted .unnumbered}

Back at the start, ChatGPT was dated to a portion of the internet from September 2021 [@azamfirei2023large]. It cannot access the internet and so you will not get anything more recent. Whereas Bard is trained from Google search results, Wikipedia, [The Common Crawl](commoncrawl.org) and other publicly available datasets. This has now changed, but it may well be that the AI you use has not accessed the most recent research. Currently, you are a much better researcher and judge of the written work than any AI. So, don't believe what it tells you without checking everything!

@@

## Will others detect ChatGPT if I use it?
Some people claim that they can detect the output from ChatGPT as it lacks the depth and insight that original authors usually have. In other words, the aim of ChatGPT is to produce the words (in a grammatically correct manner), while the aim of an author is to transmit an idea to the reader. As you might expect and hope, AI is not at the point where it can generate the same intent to communicate.

### 2024-09-11 2.16-foreign.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

Reading this book, you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work. Be aware also that biological sciences is biased in where studies take place, and who publishes them [@culumber2019widespread; @nunez2021monolingual]. What we really need then are more studies from less studied parts of the world, and we need these to be researched and published by people from those countries. Why then do we insist that science be written in English?

@@

One of my 'pet hates' is that many non-English speaking editors seem to insist that you have your work checked by a 'native English speaker' (see Figure \@ref(fig:editor)). In my experience, having advice from a 'native English speaker' is no guarantee to getting a well-written chapter or manuscript. Most of the world’s scientists did not grow up speaking English. Yet, rightly or wrongly, English is the language in which science is currently written. So, if English isn’t your mother tongue, should you expect to receive help during peer review?

@@

@@

Perhaps I should start from the outset by stating that English is my mother tongue, and that I have spent many hours correcting the language of colleagues for whom it wasn’t. However, most of these hours were spent when I was a postgraduate student or postdoc. I no longer think that correcting English is my role either as an editor, reviewer or as an advisor.

@@

@@

Failing that, I’m afraid that the best route will be to pay for help.

>>>>

Reading this book, you will learn of some biases in biological sciences and how you need to be aware of them in order to properly interpret your work. Be aware also that biological sciences is biased in where studies take place [among [a lot of other biases](https://howtopublishscience.org/pubbias.html): @measey2022how], and who publishes them [@culumber2019widespread; @nunez2021monolingual; @martin2012mapping]. What we really need then are more studies from less studied parts of the world, and we need these to be researched and published by people from those countries. Why then do we insist that science be written in English?

@@

## Do I need to write in perfect English?

The best news is that the plethora of translation tools from various languages to English is getting so good that although learning how to read and write good English is always going to be useful, it is not essential. I would suggest that instead, you put your efforts into learning scientific writing style (as taught in this book and others) instead of learning the rules of English grammar. What we are interested in is being able to communicate your science to a global (mostly non-English speaking) audience. There is a world of difference between being able to write good English and being able to write scientific English, and most English speakers struggle to make this change unless they are taught. The aim of this book is to teach people how to change their style into scientific English.

One of the major difficulties for non-English speaking students learning to write scientific English is that they don't appreciate the difference between good English and scientific English. Actually, you don't need to learn to write English, you can learn to write in the scientific version of your own language. But that must encompass the same scientific style that is being taught in this book. If you want to learn what the scientific style reads like in your own language, then I suggest making a translation of a paper in a journal that forms part of your [core target reading](#critical). Then examine this translated paper using the [critical reading](#critical) tools you'll find later in [Part II](#critical).

If you are not confident in writing in English, then write in your own language and have it translated. But you must make sure that the style used is the correct scientific style.

As a reviewer, I read plenty of manuscripts submitted by people who are writing English as a second (third or more) language, or writing in their own language and having it translated. Those that have learned the scientific style generally do very well with minor issues in grammar. They will not get penalised for these minor grammatical issues, even though they could clear most of these up with a [good grammar editor](https://www.grammarly.com/). Those who don't produce manuscripts are not suffering with language as much as a failure to understand the scientific style.

Specifically:

 - Their [citations](#citations) are poor or incorrect
 - Their studies often lack a [hypothesis](#hyp) or [clear aim](#nohyp)
 - Their [arguments](#argument) are poorly made or inconsistent
 - The study approach or study design is unclear
 - The writing is not [concise](#concise)

The good news about all of these deficiencies is that it is possible to learn how to do these well in your own language, and that none of them really have anything to do with English, but everything to do with writing in a good scientific style. Solving these problems can be done in combination. First, take a good guide to scientific writing ([like this one](#contribute)) and try to incorporate the lessons taught into your writing. Second is [critical reading](#critical) which should tell you all you need to know about how to write well. My suggestion is to do both. The book can help you to demystify the scientific style, and [critical reading](#critical) is a great skill to acquire and will take you forward with all you need to learn. Lastly, and perhaps most important, is to put into practice what you learn in your own writing. Don't let having to submit in English hold you back. Move forwards in your own language and apply all of the tools you learn and then translate once you've finished. As time goes by and your English improves, you'll learn that it's really not as hard as it seems.

@@

One of my 'pet hates' is that many non-English speaking editors seem to insist that you have your work checked by a 'native English speaker' (see Figure \@ref(fig:editor)). In my experience, having advice from a 'native English speaker' is no guarantee to getting a well-written chapter or manuscript. I have had many native English speaking students who are not able to spot very obvious mistakes, and make many of their own. Most of the world’s scientists did not grow up speaking English. Yet, rightly or wrongly, English is the language in which science is currently written. So, if English isn’t your mother tongue, should you expect to receive help during peer review?

@@

Perhaps I should make it clear that English is my mother tongue, and that I have spent many hours correcting the language of colleagues for whom it wasn’t. However, most of these hours were spent when I was a postgraduate student or postdoc. I no longer think that correcting English is my role either as an editor, reviewer or as an advisor.

@@

First, I would suggest that you check out the wide array of tools available online. Some may be behind paywalls, but provide you with a free trial. If the service that they offer is good, then consider subscribing so that they can help you in future.

Your university may have a facility to help you with English language translations.

@@

Failing that, the best route will likely be to pay for help. There are many services available to help you, and many scientific publishers now promote these if you are submitting a manuscript.

### 2024-09-11 2.17-plagiarise.Rmd
Add files via upload
# Changes to the book since April 2024 with several new chapters & core-skills as Part 1

Plagiarism is when you copy somebody else's work. We're most familiar with plagiarism of writing and these days this is especially easy with the copy and paste function. Most people are not even aware that they have plagiarised because sometime in the past they copied and pasted the work of somebody else into a document and this later became incorporated into their text without them being aware of it.

Plagiarism is a problem because, essentially, you are taking somebody else's work without attributing it to them.

@@

## Free online plagiarism tools

@@

- Read the sentence that has been plagiarised several times to yourself. 
- Now without looking at that sentence, write another sentence that has the same meaning. Because it's very hard to replicate somebody else's words without copying them, what you should find is that you've written a fresh unplagiarised sentence. 
- This can now be added to your text, changed as appropriate to fit your existing text. And that should be the end of your plagiarism worries.

@@

## How can you make sure that you never plagiarise?
Quite simply, if you never cut and paste, you won’t plagiarise. It’s that easy. I understand why people copy and paste as a way to get started, or because someone else has written something so well, it’s hard to believe that you could ever write it any better. But actually, you can write it just as well, and writing it in your own words is worth so much more.

Don’t forget that the penalty for plagiarism in your thesis might well be that you fail.

@@

Want to read more about plagiarism to understand what it is: see Louw [-@louw2017defining]

## What is autoplagiarism?

### Developmental recycling {#develop}

### Generative recycling {#generative}

### Adaptive recycling

### Duplicate recycling

## Overall - just don’t plagiarise

The simple way around this problem is not to plagiarise any of your own text (self-plagiarism) or any of anybody else's. This is, as I've explained above, relatively easy because as long as you don't copy and paste any text you'll find it very hard to actually use somebody else's words. However it can get a little bit tricky when you are writing the materials and methods section of a paper, and especially when that paper uses exactly the same materials and methods as a previous paper. The temptation is great to go and copy the text that you've written before. But be aware that, however grey, this is plagiarism and you should not do it.

The likelihood is that because you're using the same names and their relationship is staying fixed as shown in the example above in materials and methods you are more likely to come up with text that a piece of software like turn it in will show is plagiarised even when it isn't.

>>>>

Plagiarism is when you copy somebody else's work. We're most familiar with plagiarism of writing and these days this is especially easy with the copy and paste function. Whether the source is the internet, a published (or unpublished) paper, or a translation of something written in another language, if you copy someone else's work and paste it into your own this is blatant plagiarism.

When finding that there is plagiarism, I find that most people are not even aware that they have plagiarised. I think that this is because at some point in the past they copied and pasted the work of somebody else into a document and this later became incorporated into their text without them being aware of it. If this is your style of writing, then I suggest that you move away from copy and paste mistakes as it can land you in a lot of trouble. Rather make your own notes.

Plagiarism is a problem because, essentially, you are taking somebody else's work without attributing it to them. However, blatant plagiarism is only one of several kinds of plagiarism, and it is important that you know all of the different kinds of plagiarism so that you know how to avoid it. Studies of postgraduate students have found that although all are familiar with blatant plagiarism, more subtle plagiarism is not understood by all [@shen2021chinese].

It is really easy to avoid plagiarism. Follow these three easy steps (Table \@ref(tab:plagiarise)).

Table: (\#tab:plagiarise) **Three simple steps to avoid plagiarism.** Follow these three simple steps to avoid plagiarism in your work.

|  Step 1 | Step 2 | Step 3 |
| ----- | ---- | ---- |
| Read the text written by someone else from which you want to draw inspiration from | Make basic (not detailed) notes on this written work | Put the original work out of sight, and using only your notes write out your own work |

There is one final **Step 4** to the above Table \@ref(tab:plagiarise), and that is whether or not to [cite](#citations) the work that you have just used. Citations are the topic of another chapter (see [here](#citations)), and you should be familiar with what constitutes a citation and what doesn't. If you are taking an idea that is first provided in this specific text, and you are re-writing that concept, then you will need to cite the original work. If you are using someone else's writing as an inspiration for your own work, then it may or may not require a citation, and you will need to read the chapter on citations (see [here](#citations)) in order to judge.

Let's try looking at an example:

**Example: Darwin's survival of the fittest**

> "It is not the strongest of the species that survives, nor the most intelligent that survives. It is the one that is the most adaptable to change, that lives within the means available and works co-operatively against common threats." ~ Charles Darwin [-@darwin1859origin]

If you want to use Darwin's idea, then you could re-write the concept above, but then you would need to cite Darwin for this idea. As follows:

**My text:** Darwin [-@darwin1859origin] stressed that it is individuals that are most adaptable to change who will survive, rather than the strongest or most intelligent of the species.

But let us suppose that you came across this idea not from reading the original text of Darwin [-@darwin1859origin], but from another author who was writing about the same concept.

> "Darwin proposed that the capacity of organisms to produce more offspring that can be supported by the environment would lead to a struggle for existence, and individuals that are most fit for survival and reproduction would be selected through natural selection."  ~ Samuel A. Cushman [-@cushman2023entropy]

If this was the text that inspired you to use the concept, you should not cite Cushman [-@cushman2023entropy] for the idea, nor should you use any of his text. Instead, I'd suggest that you go back to Darwin [-@darwin1859origin], and take your inspiration from there.

To take this to another step, I asked [ChatGPT](https://chat.openai.com/) to rewrite the text of Chushman [-@cushman2023entropy].

**ChatGPT text:** Darwin suggested that when organisms produce more offspring than their environment can sustain, it triggers a competition for survival. Through natural selection, the individuals best adapted for survival and reproduction are favored.

While the text provided by ChatGPT is not wholly incorrect, there are some subtleties that are worth inspecting. For example, the term "survival of the fittest" was coined by Herbert Spencer [-@spencer1882principles], not Charles Darwin and there is some consensus that Spencer's [-@spencer1882principles] use of this term was not completely aligned with that given by Darwin [@offer2014natural]. It appears that Chushman [-@cushman2023entropy] linked Spencer's term to Darwin's [-@darwin1859origin] concept by using the word "fit", which is commonly the case. However, ChatGPT uses "best adapted for survival and reproduction" which is our modern definition for fitness, but not what Darwin [-@darwin1859origin] originally wrote: "the strongest... [and] ...the most intelligent". The point here is that when you cite the original text, you should go back to this text, and seek out what the author wrote and intended, not a later citation or interpretation.

Next, when you are wanting to cite an original text, you are often interested in the variables that have been investigated in that written work (see [Part II](#hypothesis)). In the example of Darwin (above) this would be the variables strength and intelligence, and if your variable was reproductive fitness you would need to explain how this was derived from Darwin's [-@darwin1859origin] original concept. Of course, you could opt to use the original text in quotation marks and cite the text (and page) where this was from. However, this is not always appropriate, so you should get used to writing your own text about other people's findings, concepts and ideas.

The other point about the ChatGPT text is that while it does not plagiarise Darwin [-@darwin1859origin] or Chushman [-@cushman2023entropy], we do not know if it has plagiarised other text from its input data: text from the internet which is mostly derived from original input by human authors (see chapter on AI [here](#AI)). Moreover, as increasing amounts of text on the internet will be derived from AI, we have no idea how skewed and misrepresentative AI generated text will become.

There is more to plagiarism than meets the eye, and it is important that you have a good understanding of what plagiarism is as many postgraduate students only appear to understand the most blatant kind of plagarism when making a verbatim copy of someone else's text [@shen2021chinese].

## The different kinds of plagiarism

It is important to know what plagiarism is, so that you know how to avoid it. While almost everyone understands that they must not use other people's work verbatim, it is also important to know that other more subtle forms of plagiarism.

### Blatant or verbatim plagiarism

This is when you 'copy and paste' someone else's work and pretend that it is your own. Note that even if you didn't use the copy/paste functions, if you have incorporated someone else's written work into your own text, then this is plagiarism. This includes using a translated version of work in another language.

### Paraphrasing

Some students will copy an entire sentence and then change one or two words; this is known as paraphrasing and is also plagiarism. Subtle variations on this including switching the order of the clauses in a sentence and substituting some words with synonyms. The essence of paraphrasing is that you start with the verbatim form of plagiarism, and you are then attempting to engineer some differences by remixing this content. Instead you should take the three easy steps (Table \@ref(tab:plagiarise)) and write it yourself. Remember that you may need a citation (see above).

**What follows are more subtle forms of plagiarism:**

### Failure to attribute

If someone else wrote pieces of your thesis and you do not attribute their input, this is plagiarism. There are many people that could help with your writing, such as friends, lab mates, or your advisor. But if you then pass this work off as your own, it is plagiarism.

With failure to attribute plagiarism, it is not as simple as stating that some text was written by your friend or lab mate. This is not an option. When you submit your PhD, you will declare that the work is your own and that you have written all of it.

There is another form of failure to attribute plagiarism when you employ a third party to write your thesis for you. This process is often referred to as a "[paper mill](https://howtopublishscience.org/authors.html#papermills)" [@measey2022how]. Even if you paid this person and they gave you permission to use the text in your thesis, it is still plagiarism.

#### Is it plagiarism if an AI bot writes it?
I would also place the use of AI as a form of failure to attribute plagiarism. Currently, we are moving into the new territory of AI generated text. You can not use AI generated text (e.g. ChatGPT or Bard) in your PhD or in manuscripts submitted to journals as it is prohibited. Most institutions make you sign a declaration that you will never do this. You can read more on this topic in another chapter [here](#AI).

#### Avoiding failure to attribute plagairism

Avoiding failure to attribute plagiarism is just the same as avoiding blatant plagiarism (see [below](#anitplagiarise)). Don't copy and paste their work. Do read it, do make notes, but then put it out of sight and write your own work (see Table \@ref(tab:plagiarise)).

#### What is co-authored work makes up part of your thesis?
Many theses are made up of multi-authored papers. The reason for this is that different people in the author line contribute different aspects of the paper. If you are in the situation where one of your chapters is partly written by your co-author, what should you do?

The reality is that when you submit such multi-authored work as part of your thesis, you should also have an opportunity to declare who did what in the manuscript (see the CREDIT Contributor Checklist [here](#credit)). You should have written the original draft, including all of the text. While other members of the author line have have read and reviewed your work, they should not have written any of the sections themselves. If they did, then you should take follow the three easy steps (Table \@ref(tab:plagiarise)) and turn it into your own writing.

### Autoplagiarism

#### Developmental recycling {#develop}

#### Generative recycling {#generative}

#### Adaptive recycling

#### Duplicate recycling

@@

## Free online plagiarism detection tools

@@

- Read the sentence that has been plagiarised several times to yourself. 
- Make a few notes about the meaning of the text, so that you don't need to look back at it (avoid using the same words)
- Move the original text away so that you cannot see it
- Now without looking at that sentence, write another sentence that has the same meaning. Because it's very hard to replicate somebody else's words without copying them, what you should find is that you've written a fresh unplagiarised sentence. 
- This can now be added to your text, changed as appropriate to fit your existing text. And that should be the end of your plagiarism worries.

This can be summarised into three easy steps (see Table \@ref(tab:plagiarise)).

@@

## How can you make sure that you never plagiarise? {#anitplagiarise}

Quite simply, if you never cut and paste, you won’t plagiarise. It’s that easy. I understand why people copy and paste as a way to get started, or because someone else has written something so well, it’s hard to believe that you could ever write it any better. But actually, you can write it just as well, and writing it in your own words is worth so much more.

Don’t forget that the penalty for plagiarism in your thesis might well be that you fail. More than that, you could lose your place at the university or even your job [@ruiperez2016plagiarism; @tatalovic2017croatia].

@@

## In summary - just don’t plagiarise

Plagiarism is the unattributed use of other people's work. It is not simply the copying of other people's text verbatim. There are many forms, and you need to be aware of the extent of that which is considered to be plagiarism in order to avoid it.

The simple way around plagiarism is not to copy any of your own text (self-plagiarism) or any of anybody else's. This is, as I've explained above (see Table \@ref(tab:plagiarise)), relatively easy because as long as you use your own words, you should be plagiarism free. The temptation will always be great to go and copy the text that has been used before. But be aware that, however grey, this is plagiarism and you should not do it.
